---
title: "Clustering and PCA"
subtitle: "Data Science for Biologists, Spring 2020"
author: "Olivia Smithson"
date: "04/23/2020"
output: 
  html_document:
    highlight: zenburn
    toc: TRUE
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

## SET YOUR SEED BELOW!!! Make it YOUR SEED! Uncomment the line, and add your chosen seed number into the parentheses
set.seed(825)
```

## Instructions

Standard grading criteria apply. And...

+ Make sure you set your seed
+ Proofread to submit YOUR OWN WORDS!!
+ Make sure all bullets properly render in the knitted HTML. **YOU WILL LOSE POINTS FOR BULLETED LISTS THAT DO NOT TURN OUT PROPERLY IN THE HTML!!** 
+ Remember that `#` do NOT make comments in markdown. They make HEADERS!! You need to use HTML comments to have comments in your text that don't appear in the knitted HTML. For example, the line below does not appear in html.

<!-- This is an actual comment in markdown. -->

## The data

We will be using a dataset of olives collected from various regions of Italy. The data gives the percentage of different fatty acids that comprise the oil in the olives. [Source](https://rdrr.io/cran/dslabs/man/olive.html)

```{r, collapse=T}
olives <- read_csv("https://raw.githubusercontent.com/sjspielman/datascience_for_biologists/master/data/olives.csv")
dplyr::glimpse(olives)
```


## Part 1: Clustering

Perform TWO k-means cluster analysis, once with **k=3** and once with **k=5**. 

```{r, collapse=TRUE}
## Perform two clustering analyses here with k=3 and k=5, separately.

## k=3 clustering
olives %>%
  select(-region, - area) %>%
  ##specify k=3
  kmeans(3) -> olives_k3

## k=5 
olives %>%
  select(-region, -area) %>%
  ##specify k=5
  kmeans(5) -> olives_k5
```



For each analysis, make a SINGLE FIGURE for each clustering analysis (two figures total). Choose your own adventure for each figure! You should explain per figure: a) Why did you decide to visualize the clustering using this particular figure? (~1 sentence) and b) What insights (if any!) does this figure provide (1-2 sentences)? Answer in the bullets below the given figure. 

**You should not make two identical figures - they need to be _somewhat_ different.** For example, two barplots showing the relationship between clusters and regions is a no-go. Two scatterplots but with different x-axis and y-axis variables on the axes is ok! Two boxplots (violin, den sity, histogram, sina, jitter...) but with a different focal variable is ok! Either way, your figure MUST show the clusters.

```{r,collapse=TRUE}
## Figure for k=3 including all associated wrangling

##Plot #1:
olives %>%
  mutate(k3_clusters = factor(olives_k3$cluster)) %>%
  ggplot(aes(x = k3_clusters, y = palmitic, fill = k3_clusters)) +
  geom_violin() +
  labs(x = "K-means Clusters",
       y = "Palmitic Acid") +
  scale_fill_brewer(palette = "Greens") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.background = element_rect(fill = "lemonchiffon2")) 


```

+ **Answer: Why did you choose this figure?**
+ I chose to do a violin plot of palmitic acid across clusters because I always use skin care products with palmitic acid in them, and I knew the violin plot would present a clean output for clustering analysis.  
+ **Answer: What trends/insights (if any)  does the figure reveal, as they relate to the clusters?**
+ Within this violin plot, there are 3 distinct clusters visible for the palmitic acid coordinate space. Cluster 1 appears to be the most variable.


```{r, collapse=TRUE}
## Figure for k=5 including all associated wrangling
##Plot #2:
olives %>%
  mutate(k3_clusters = factor(olives_k3$cluster)) %>%
  ggplot(aes(x = region, y = eicosenoic, color = k3_clusters)) +
  geom_jitter() +
  labs(x = "Region",
       y = "Eicosenoic Acid") +
  scale_color_manual(name = "K-Means Cluster", values = c("cornflowerblue", "aquamarine4", "lightsalmon4")) +
  theme_minimal() +
  theme(plot.background = element_rect(fill = "cornsilk"))

```

+ **Answer: Why did you choose this figure?**
+ I tried to do a scatter plot for this figure because we had practiced with them in the lesson for this week, but that attempt made me realize that a jitter plot would be more appropriate with continuous and discrete variables. I also recently learned about eicosenoic acid in Biochemistry so that olive component jumped out at me. 
+ **Answer: What trends/insights (if any) does the figure reveal, as they relate to the clusters?**
+ While there are no fully distinct clusters within this jitter plot by region, it appears that Northern Italy's coordinate space is primarily filled with points from Cluster 3, Sardinia's coordinate space is primarily filled with points from Cluster 2, and Southern Italy's coordinate space is primarily filled with points from Cluster 1. Southern Italy's jitter arrangment is the most variable. 


## Part Two: PCA

Perform a principal components analysis on this dataset, and create TWO figures from your PCA. Each figure should have a brief explanation as a bullet below it. For each figure, your should be labeled to include information about how much variation the given PC explains in the data.

```{r, collapse=TRUE}
## Code to perform PCA goes here

olives %>%
  ##remove non-numeric variables
  select(-region, -area) %>%
  ##scale the data to ensure stochasticity and ensure a matrix is the output
  scale() %>%
  prcomp() -> olives_pca


```

1. But first, use code to answer this question! Intepret how variables in the initial dataset **load** on the PCs by examining, you guessed it, the loadings. In the bullets below, state which SINGLE variable load most strongly on PC1 and PC2 each, and which SINGLE variable loads most weakly on PC1 and PC2.

```{r, collapse=TRUE}
## Code to examine loadings goes here
olives_pca$rotation
##oleic acid loads strongly to PCA 1, which means it contributes the most to the information
##extra PCA notes:
##if arrows are pointed 180 degrees apart, they are oppositely-related (contributing opposite info)
##if arrows are pointed 90 degrees apart, they are related orthogonally so they might have one variable in common
```

+ **Strongest PC1 loading: oleic acid (despite it being negative)**
+ **Strongest PC2 loading: linoleic acid**
+ **Weakest PC1 loading: stearic acid**
+ **Weakest PC2 loading: palmitic acid**

2. Plot PC2 against PC1 (PC2 is on Y and PC1 is on X) as a scatterplot, where you use color to distinguish among **olive region** (NOT area!). Make sure to show ellipses for each grouping. Do NOT include loading arrows. In the bullet below, _explain in 2-3 sentences_ if and how either PC1 or PC2 discriminates olive region. You may write a bulleted list *as long as it renders properly.*

```{r, collapse=TRUE}
##Code to make this figure goes here
as_tibble(olives_pca$x) %>%
  mutate(region = olives$region) %>%
  ggplot(aes(x = PC1, y = PC2, color = region)) +
  geom_point() +
  scale_color_manual(name = "Region", values =  c("darkseagreen4", "darkslategrey", "gold4")) +
  ##the argument stat_ellipse forms an ellipse around whatever is colored in the plot. In this case, that's region.
  stat_ellipse() +
  xlab("PC1") +
  ylab("PC2") + 
  theme_minimal() +
  ##must use element_rect() with plot.background!
  theme(plot.background = element_rect(fill = "lemonchiffon2"))

##PCA uses linear algebra to create a version with transformed axes to examine variation among portions of the dataset
  
```

+ **Explanation for figure:**
+ Based on the plot, it appears that PC1 discriminates all 3 regions from each other, whereas PC2 discriminates Sardinia from Northern and Southern Italy. This is because Sardinia appears more concentrated when compared from Northern and Southern Italy despite their similar locations on the PC2 axis.

2. Plot PC2 against PC1 as a scatterplot, where you use color to distinguish among **olive area** (the other categorical column!). Make sure to show ellipses for each grouping. Do NOT include loading arrows. In the bullet below, _explain in 2-3 sentences_ if and how either PC1 or PC2 discriminates olive region. You may write a bulleted list *as long as it renders properly.*

```{r, collapse=TRUE}
as_tibble(olives_pca$x) %>%
  mutate(area = olives$area) %>%
  ##for scatterplot, x and y are PCAs
  ggplot(aes(x = PC1, y = PC2, color = area)) +
  geom_point() +
  ##easier to use a brewer palette here
  scale_color_brewer(palette = "BrBG") +
  stat_ellipse() +
  theme_minimal() +
  ##only fill that works for the background here is white
  theme(plot.background = element_rect(fill = "white"))

##Sardinia's are next to each other 

```

+ **Explanation for figure:**
+ Both PC1 and PC2 do not discriminate Coast and Inland Sardinia. PC1 does not discriminate East and West Liguria, Umbria, or North-Apulia from each other, but PC2 discriminates all four from each other. North and South Apulia are discriminated from each other on both axes.


## BONUS QUESTION 1

Make a figure of the loadings on PC1 and PC2 in the space below. Make sure it is as legible as possible! No explanation needed, just make a plot!
```{r, collapse=TRUE, fig.height=6, fig.width=8}
## Bonus loadings plot

##Wrangling (slide 13 is helpful for this)
##convert to data frame and then tibble because it starts out as a matrix
olives_pca$rotation %>%
  as.data.frame() %>%
  rownames_to_column("quantity") %>%
  as_tibble() -> olives_pca_loadings

##create arrow specifics
dr_spielman_arrow <- grid::arrow(length = unit(0.05, "inches"), type = "closed") 

##create PC1 and PC2 loadings figure
olives_pca_loadings %>%
  ggplot(aes(x = PC1, y = PC2)) +
  ##specify arrow in geom_segment
  geom_segment(x = 0, 
               y = 0, 
               arrow = dr_spielman_arrow, 
               aes(xend = PC1, yend = PC2)) +
  ##play around with text to make it suitable
  geom_text(aes(label = quantity),
                 size = 4, color = "darkolivegreen4") +
  theme_minimal() +
  theme(plot.background = element_rect(fill = "lemonchiffon2"))
##may have to set xlim and ylim depending on the output for future plots



```

## BONUS QUESTION 2

Perform a k-means clustering with *k=3* on the PCs to examine whether olive *regions* cluster together. **You must include a clear explanation (1-3 sentences) to receive ANY CREDIT AT ALL for this bonus.**
```{r}
## Bonus clustering
```
+ **Bonus explanation**

